
####Marijuana, Alcohol, and Other Drugs  

**Continuing our Exploration of the Data**
Now having looked a bit deeper into the 

**What we'll do in this post**

* Load the csv, format the data
  * This will all be hidden and can be found in the previous post (Part 1)
* Look into apparent growth in crime rates from 2012 - 2014
* We'll focus only on those that fit the "IS_CRIME" definition and not "IS_TRAFFIC"

Let's dive in!  

**Exploration of Data**  
Data provided by http://data.denvergov.org

**Import the necessary libraries**

```{r results='hide', warning=FALSE, message=FALSE, echo=FALSE}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(ggmap)
#options("stringsAsFactors" = TRUE)
```  

```{r cache=FALSE, echo=FALSE}
####
# Data from: http://data.denvergov.org/dataset/city-and-county-of-denver-crime
# File name: crime.csv
CWD = getwd()
data = read.csv(paste(CWD,'/data/crime.csv',sep=''))
data.backup=data
####
```  

```{r echo=FALSE}
#data=data.backup
#Format FIRST_OCCURRENCE_DATE as.Date and use as crime date (for now)
data$date = as.Date(data$FIRST_OCCURRENCE_DATE)
#Create new columns for grouping
data$year = year(data$date)
data$month = month(data$date)
data$day = day(data$date)
data$dayOfWeek <- weekdays(data$date)
data$hour = hour(data$FIRST_OCCURRENCE_DATE)
data$dayOfWeek = factor(data$dayOfWeek, levels= c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
maxYear = max(data$year)
maxMonthYTD = max(data$month[data$year==maxYear])

data$DRUG = 'OTHER'
data$DRUG[grepl('marijuana',data$OFFENSE_TYPE_ID)] = 'MARIJUANA'
data$DRUG[grepl('alcohol',data$OFFENSE_TYPE_ID)] = 'ALCOHOL'
data$DRUG[grepl('liquor',data$OFFENSE_TYPE_ID)] = 'ALCOHOL'

data$CRIME_TYPE = 'OTHER'
data$CRIME_TYPE[grepl('possess',data$OFFENSE_TYPE_ID)] = 'POSSESS'
data$CRIME_TYPE[grepl('sell',data$OFFENSE_TYPE_ID)] = 'SELL'

data$incidents = 1

```  

**Looking Into Drugs and Alcohol**  

```{r}
top10df = data %>%
  group_by(OFFENSE_TYPE_ID) %>%
  filter(OFFENSE_CATEGORY_ID=='drug-alcohol')%>%
  summarise(incidents = sum(IS_CRIME) + sum(IS_TRAFFIC)) %>%
  arrange(desc(incidents))

top10 = top15df$OFFENSE_TYPE_ID[1:10]

df = data %>%
  group_by(year,OFFENSE_TYPE_ID) %>%
  filter(OFFENSE_TYPE_ID %in% top15)%>%
  filter(month < maxMonthYTD) %>% 
  summarise(incidents = sum(IS_CRIME) + sum(IS_TRAFFIC)) %>%
  arrange(desc(incidents))

p = ggplot(df, aes(x=reorder(OFFENSE_TYPE_ID,incidents),y=incidents,fill=factor(year),col=factor(year)))
p + geom_bar(stat='identity') + coord_flip()

df$DRUG = 'OTHER'
df$DRUG[grepl('marijuana',df$OFFENSE_TYPE_ID)] = 'MARIJUANA'
df$DRUG[grepl('alcohol',df$OFFENSE_TYPE_ID)] = 'ALCOHOL'
df$DRUG[grepl('liquor',df$OFFENSE_TYPE_ID)] = 'ALCOHOL'

df$CRIME_TYPE = 'OTHER'
df$CRIME_TYPE[grepl('possess',df$OFFENSE_TYPE_ID)] = 'POSSESS'
df$CRIME_TYPE[grepl('sell',df$OFFENSE_TYPE_ID)] = 'SELL'

df2 = df[df$CRIME_TYPE != 'OTHER',]

p = ggplot(df2,aes(x=factor(year),y=incidents,col=OFFENSE_TYPE_ID,fill=OFFENSE_TYPE_ID))
p + geom_bar(stat='identity') + facet_wrap(~OFFENSE_TYPE_ID) + theme(legend.position="none")



#p = ggplot(df, aes(factor(dayOfWeek),crime,fill=factor(year),col=factor(year)))
#p + geom_boxplot()

#p = ggplot(df, aes(factor(dayOfWeek),traffic,fill=factor(year),col=factor(year)))
#p + geom_boxplot()

```

```{r}
#head(data)
df = data %>%
  filter(DRUG != 'OTHER') %>%
  filter(CRIME_TYPE != 'OTHER') %>%
  filter(month < maxMonthYTD) %>% 
  spread(CRIME_TYPE,incidents)
df[is.na(df)] = 0

df2 = df %>%
  filter(year>=2014) %>%
  group_by(date) %>%
  summarise(possess=sum(POSSESS),sell=sum(SELL)) %>%
  filter(sell > 0)
df2[is.na(df2)] = 0

head(df2)
p = ggplot(df2,aes(x=sell,y=possess,col=factor(year(date))))
p + geom_point()


df = data %>%
  filter(DRUG != 'OTHER') %>%
  filter(CRIME_TYPE != 'OTHER') %>%
  filter(year >= 2013) %>%
  group_by(year,month,CRIME_TYPE) %>%
  summarise(incidents=sum(incidents))
df$yearmonth = paste(df$year,df$month, sep="-")
p = ggplot(df,aes(x=factor(year),y=incidents,col=CRIME_TYPE))
p + geom_bar(stat='identity')



```


```{r}

denver = get_map('denver',zoom=12)
denverMap = ggmap(denver)
denverMap +
stat_density2d(
aes(x = GEO_LON, y = GEO_LAT, fill = ..level..,
alpha = ..level..),
size = 2, bins = 5, data = data %>% filter(IS_CRIME==1) %>% group_by(GEO_LAT,GEO_LON) %>% summarise(incidents=sum(IS_CRIME)),
geom = "polygon")

```

```{r}

df = data %>% filter(IS_CRIME == 1) %>% group_by(GEO_LON,GEO_LAT,year) %>% summarise(incidents=sum(IS_CRIME))

denver = get_map('denver',zoom=13,color='bw')
denverMap = ggmap(denver,base_layer=ggplot(aes(x=GEO_LON,y=GEO_LAT),data=df))

denverMap + stat_density2d(aes(x=GEO_LON,y=GEO_LAT, fill = ..level.. , alpha = ..level..), bins=5,geom='polygon',data=df) + scale_fill_gradient(low='black',high='red') + facet_wrap(~year)

```
```{r}
#head(data)

df = data %>%
  filter(OFFENSE_TYPE_ID=='traffic-accident-dui-duid') %>% 
  group_by(GEO_LON,GEO_LAT,hour) %>% 
  summarise(incidents = sum(IS_TRAFFIC))

#denver = get_map('denver',zoom=13,color='bw')
#denverMap = ggmap(denver,base_layer=ggplot(aes(x=GEO_LON,y=GEO_LAT),data=df))

#denverMap + stat_density2d(aes(x=GEO_LON,y=GEO_LAT, fill = ..level.. , alpha = ..level..), bins=5,geom='polygon',data=df) + scale_fill_gradient(low='black',high='red') + facet_wrap(~hour)


```


```{r}
df = data %>%
  filter(OFFENSE_TYPE_ID=='traffic-accident-dui-duid') %>% 
  group_by(date,dayOfWeek,hour) %>% 
  summarise(incidents = sum(IS_TRAFFIC))

p = ggplot(df,aes(x=factor(hour),y=incidents,fill=dayOfWeek,col=dayOfWeek))
p + geom_boxplot()

df = data %>%
  filter(OFFENSE_TYPE_ID=='traffic-accident-dui-duid') %>% 
  group_by(date) %>% summarise(incidents = sum(IS_TRAFFIC))
ggplot(df,aes(x=date,y=incidents)) + geom_line()       
unique(data$OFFENSE_TYPE_ID)
```



```{r}
df = data %>% filter(IS_CRIME == 2) %>% filter(filter(IS_OFFENSE_TYPE=='')) %>% group_by(GEO_LON,GEO_LAT,year) %>% summarise(incidents=sum(as.numeric(incidents)))

denverMap = ggmap(denver,base_layer=ggplot(aes(x=GEO_LON,y=GEO_LAT),data=df))

denverMap + stat_density2d(aes(x=GEO_LON,y=GEO_LAT, fill = ..level.. , alpha = ..level..), bins=5,geom='polygon',data=df) + scale_fill_gradient(low='black',high='red') + facet_wrap(~year)
```


```{r}
df = data %>% 
  filter(IS_CRIME==1) %>%
  filter(OFFENSE_CATEGORY_ID == 'all-other-crimes') %>%
  filter(month < maxMonthYTD) %>%
  group_by(OFFENSE_TYPE_ID,year) %>%
  summarise(incidents=sum(IS_CRIME)) %>%
  arrange(desc(incidents))

p = ggplot(df,aes(x=reorder(OFFENSE_TYPE_ID,incidents),y=incidents,fill=factor(year),col=factor(year)))
p + geom_bar(stat='identity') + coord_flip()
```

```{r}
df = data %>%
  filter(IS_CRIME==1) %>%
  group_by(OFFENSE_TYPE_ID) %>%
  summarise(incidents=sum(IS_CRIME)) %>%
  arrange(desc(incidents))

top20 = df$OFFENSE_TYPE_ID[1:20]

df = data %>%
  filter(OFFENSE_TYPE_ID %in% top20) %>%
  group_by(OFFENSE_TYPE_ID,dayOfWeek) %>%
  summarise(incidents=sum(IS_CRIME))

ggplot(df,aes(dayOfWeek,OFFENSE_TYPE_ID)) + geom_tile(aes(fill=incidents),col='white') + scale_fill_gradient(low = "white", high = "steelblue")
```

```{r}
df = data %>% 
  filter(IS_TRAFFIC==1) %>%
  filter(month < maxMonthYTD) %>%
  filter(OFFENSE_TYPE_ID == 'traffic-accident-dui-duid') %>%
  group_by(dayOfWeek,date,year) %>%
  summarise(incidents=sum(IS_TRAFFIC)) %>%
  arrange(dayOfWeek)
p = ggplot(df, aes(factor(dayOfWeek),incidents,fill=factor(year),col=factor(year)))
p + geom_boxplot()
```
```{r}
df = data %>%
  filter(IS_TRAFFIC==1) %>%
  group_by(OFFENSE_TYPE_ID) %>%
  summarise(incidents=sum(IS_TRAFFIC)) %>%
  arrange(desc(incidents))

top20 = df$OFFENSE_TYPE_ID[1:20]

df = data %>%
  filter(OFFENSE_TYPE_ID %in% top20) %>%
  group_by(OFFENSE_TYPE_ID,dayOfWeek) %>%
  summarise(incidents=sum(IS_TRAFFIC))

ggplot(df,aes(dayOfWeek,OFFENSE_TYPE_ID)) + geom_tile(aes(fill=incidents),col='white') + scale_fill_gradient(low = "white", high = "steelblue")
```






**Observations**  

* "all-other-crimes" is the outright leader in change in both volume and percentage growth year-over-year with an astonishing 380% increase between 2012 and 2013
* "drug-alcohol" grew by 173% between 2012 and 2013 but dropped down to only 27% growth the next year
* "murder" didn't change too much in volume compared to everything else (swinging up 7 and down 10) but was a 19% growth and a 23% decline in 2013 and 2014 respectively  


**Final Thoughts (for now)**
Due to the vague nature of the types of crimes which grew the most, I can't determine exactly what happened in Denver during 2013. In the less vague crimes, "drug-alcohol" saw the largest increase. This was followed by "public-disorder" and perhaps there's a relationship there. My assumption is that one may perhaps cause the other...  

I'm still curious about the seasonality and month-to-month effects. Perhaps certain types of crimes are more common during certain times. I'm also very interested to see if a new population was perhaps added to the mix in 2013. If a certain part of Denver was added in 2013 that would certainly help to explain the situation.  

**What I'll do in the next crime posts**  

* Look for patterns by location
* Lay out some visualizations on maps
* Try to identify areas with high volumes of traffic incidents (maybe I can avoid a ticket)
* Answer the question: What types of crimes have grown the most in the last 5 years?

My code for this is posted on GitHub:  
https://github.com/stoltzmaniac/Denver-Crime-Analysis.git